---
title: "flux-analysis"
author: "Gavin McNicol"
date: "2026-06-03"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Purpose: Analysis of Nachusa 2022 flux data for Frontier in Environmental Sciences special issue.

# Load libraries

```{r packages}
library(tidyverse)
library(lme4)
library(rpart)
library(rpart.plot)
```

# Data

```{r read-data}
data <- read_csv("data/2022_full_data.csv") %>% 
  select(-estimate, -eco)
head(data)
```

check on groupings

```{r}
data %>% distinct(ecosystem)
```

BP = big prairie (open prairie 1; Pr-1)
SP = small prairie (open prairie 2; Pr-2)
NW = north wetlands (sloping open prairie; Wt-1)
SW = south wetlands (wet prairie; Wt-2)
EW = east woodlands (oak woodlands 1; Wd-1)
WW = west woodlands (oak woodlands 2; Wd-2)


# Wrangling

- Change ecosystems to oak woodland, open prairie, wet prairie

```{r wrangling}
data <- data %>% 
  mutate(cover = 
  case_when(
    ecosystem %in% c("BP", "SP") ~ "Pr",
    ecosystem %in% c("EW", "WW") ~ "Wd", 
    ecosystem %in% c("NW", "SW") ~ "Wt",
    TRUE ~ NA
  )) %>% 
  mutate(site = 
  case_when(
    ecosystem == "BP" ~ 1,
    ecosystem == "SP" ~ 2,
    ecosystem == "NW" ~ 1,
    ecosystem == "SW" ~ 2,
    ecosystem == "EW" ~ 1,
    ecosystem == "WW" ~ 2,
    TRUE ~ NA
  )) %>% 
  select(gas, date, cover, site, transect, rep, flux, p.value, soil_moisture, soil_temperature)
```

# Flux QC 

If `p.value` < 0.05, set flux to `NA` for CO2, and zero for CH4/N2O (bad p-value for CO2 is likely bad flux overall)
Set very high woodland CO2 emissions (> 9) related to ant burrow to `NA`

```{r flux-qc}
data <- data %>% 
  mutate(flux = 
           case_when(
             p.value > 0.05 & gas == "co2" ~ NA,
             p.value > 0.05 & gas %in% c("ch4", "n2o") ~ 0,
             gas == "co2" & flux > 9 ~ NA,
             TRUE ~ flux
           ))
```

# Add coords to create final tidy output dataset

```{r}
coords <- read_csv("2022-site-coords.csv")
data %>%
  mutate(site = as.character(site)) %>% 
  unite(site, cover, site, sep = "-") %>% 
  left_join(coords) %>% 
  write_csv("2022-nachusa-fluxes.csv")
```

 

# Data Summary

# Overall stats

```{r num-obs}
data %>% 
  filter(!is.na(flux)) %>% 
  count(gas)
```


```{r flux-dist, fig.width=8, fig.height=3}
data %>% 
  mutate(gas = fct_relevel(gas, "co2", "ch4", "n2o")) %>% 
  ggplot(aes(flux)) +
  geom_density(alpha = 0.6) +
  facet_wrap(~gas, 
             scales = "free",
             labeller = as_labeller(c(
    co2 = "CO[2]",
    ch4 = "CH[4]",
    n2o = "N[2]*O"
  ), label_parsed)) +
  theme_classic() +
  labs(x = expression(atop("GHG Flux", "("*mu*"mol CO"[2]*" m"^{-2}*" s"^{-1}*"), (nmol CH"[4]*" | N"[2]*"O m"^{-2}*" s"^{-1}*")")),
       y = "Density") +
  scale_fill_viridis_d() %>% 
  theme(
    strip.background = element_blank(),
    strip.text = element_text(
    size = 12,
    face = "bold"
  ))
ggsave("paper-figures/figS1.png", height = 3, width = 8)
```

CO2 unimodal with long right tail
CH4 bimodal with long left tail and many zeros
N2O unimodal normal around zero

```{r overall-stats}
data %>% 
  group_by(gas) %>% 
    summarize(min = min(flux, na.rm = T),
              q25 = quantile(flux, 0.25, na.rm = T),
              median = median(flux, na.rm = T),
              q75 = quantile(flux, 0.75, na.rm = T),
              max = max(flux, na.rm = T))
```

# By-cover stats

```{r num-obs}
data %>% 
  filter(gas %in% c("co2", "n2o"),
         !is.na(flux)) %>% 
  group_by(gas, cover) %>% 
  count(gas)
```

```{r num-obs}
data %>% 
  filter(gas %in% c("ch4"),
         !is.na(flux)) %>% 
  group_by(gas, cover, site) %>% 
  count(gas)
```

## Q1b. What is the center and spread of each flux, for each cover type, over the growing season?
- data were not normally distributed, except for N2O
- therefore, report min, q25, median, q75, max 
- mean and standard deviation for N2O
- by-site for CH4, aggregated for CO2 and N2O 

```{r flux-dist, fig.width=8, fig.height=3}
data %>% 
  mutate(gas = fct_relevel(gas, "co2", "ch4", "n2o")) %>% 
  ggplot(aes(flux, fill = cover)) +
  geom_density(alpha = 0.6) +
  facet_wrap(~gas, 
             scales = "free",
             labeller = as_labeller(c(
    co2 = "CO[2]",
    ch4 = "CH[4]",
    n2o = "N[2]*O"
  ), label_parsed)) +
  theme_classic() +
  labs(x = expression(atop("GHG Flux", "("*mu*"mol CO"[2]*" m"^{-2}*" s"^{-1}*"), (nmol CH"[4]*" | N"[2]*"O m"^{-2}*" s"^{-1}*")")),
       y = "Density",
       fill = "Cover") +
  scale_fill_viridis_d() %>% 
  theme(
    strip.background = element_blank(),
    strip.text = element_text(
    size = 12,
    face = "bold"
  ))
ggsave("paper-figures/fig1.png", height = 3, width = 8)
```

CO2 fluxes for woodlands are unimodal and peak at 2-3 (smallest source)
  (note, identified and removed outlier co2 flux in woodlands > 9)
CO2 flues for wetlands and prairie are bimodal with peaks at 1.5 and 4 (larger sources)

CH4 fluxes for woodlands are bimodal with peaks at -1.5 and 0 (largest sink)
CH4 fluxes for wetlands are bimodal with peaks at 1 and 0 (medium sink)
CH4 fluxes for prairie are bimodal with peaks at 0.6 and 9 (weakest sink)

N2O fluxes for all cover types are unimodal around zero, with slightly positive skew, esp for prairie

```{r}
data %>% 
  filter(gas %in% c("co2", "n2o")) %>% 
  group_by(gas, cover) %>% 
    summarize(min = min(flux, na.rm = T),
              q25 = quantile(flux, 0.25, na.rm = T),
              median = median(flux, na.rm = T),
              q75 = quantile(flux, 0.75, na.rm = T),
              max = max(flux, na.rm = T))
```

```{r}
data %>% 
  filter(gas %in% c("ch4")) %>% 
  group_by(gas, cover, site) %>% 
    summarize(min = min(flux, na.rm = T),
              q25 = quantile(flux, 0.25, na.rm = T),
              median = median(flux, na.rm = T),
              q75 = quantile(flux, 0.75, na.rm = T),
              max = max(flux, na.rm = T))
```


# Site effects

Q: is there a significant site effect within cover types?

```{r}
data %>% 
  filter(gas == "co2") %>% 
  lm(flux ~ cover+factor(site), data = .) %>% 
  summary()
```

```{r}
data %>% 
  filter(gas == "ch4") %>% 
  lm(flux ~ cover+factor(site), data = .) %>% 
  summary()
```
```{r}
data %>% 
  filter(gas == "n2o") %>% 
  lm(flux ~ cover+factor(site), data = .) %>% 
  summary()
```

# Q2 - Spatiotemporal models with moisture and tempertaure

```{r temp-moisture-pattern}
data %>% 
  filter(!is.na(date)) %>% 
  select(date, cover, soil_moisture, soil_temperature) %>% 
  pivot_longer(3:4, names_to = "soil_var") %>% 
  mutate(soil_var = fct_relevel(soil_var, "soil_temperature", "soil_moisture")) %>% 
  ggplot(aes(x = factor(date), y = value, color = cover)) +
  geom_boxplot() +
  facet_wrap(~soil_var, 
             scales = "free",
             labeller = as_labeller(c(
    soil_moisture = "Soil Moisture",
    soil_temperature = "Soil Temperature"
  ))) +
  theme_classic() +
  labs(y = expression(atop("Temp. ("*degree*"C)", "Moisture (%)")),
       x = "Sampling Date (monthly)",
       color = "Cover") +
  scale_fill_viridis_d() %>% 
  theme(
    strip.background = element_blank(),
    strip.text = element_text(
    size = 12,
    face = "bold"
  ))
ggsave("paper-figures/fig2.png", height = 3, width = 8)
```
# Soil temp and moisture summary table (Table S2)


```{r}
data %>%  
  filter(!is.na(date)) %>% 
  select(date, cover, soil_moisture, soil_temperature) %>% 
  group_by(date) %>% 
  summarize(mean_soil_t = mean(soil_temperature, na.rm = T),
            sd_soil_t = sd(soil_temperature, na.rm = T),
            mean_soil_moist = mean(soil_moisture, na.rm = T),
            sd_soil_moist = sd(soil_moisture, na.rm = T)) 
```


```{r}
data %>%  
  filter(!is.na(date)) %>% 
  select(date, cover, soil_moisture, soil_temperature) %>% 
  group_by(date, cover) %>% 
  summarize(mean_soil_t = mean(soil_temperature, na.rm = T),
            sd_soil_t = sd(soil_temperature, na.rm = T),
            mean_soil_moist = mean(soil_moisture, na.rm = T),
            sd_soil_moist = sd(soil_moisture, na.rm = T)) 
  
```

## Temperate ranges within 10-25 C

```{r}
data %>% 
  count(gas == "co2" & soil_temperature < 10 | soil_temperature > 25) 
55/593*100
```

## Temperature ranges for woodlands

```{r}
data %>% 
    filter(gas == "co2" & soil_temperature > 25 & cover == "Wd") %>% 
  summarize(count = n(),
            mean_soil_t = mean(soil_temperature),
            mean_soil_m = mean(soil_moisture))
```

## Highest temp-moisture combination in wet prairie

```{r}
data %>% 
  filter(gas == "co2" & date == "2022-08-18") %>% 
  group_by(cover) %>% 
  summarize(soil_t = max(soil_temperature, na.rm = T),
            soil_m = max(soil_moisture, na.rm = T))
``` 


```{r fig.height = 4.5, fig.width = 11}
data %>% 
  mutate(gas = fct_relevel(gas, "co2", "ch4", "n2o")) %>% 
  ggplot(aes(x = soil_temperature, y = flux, shape = cover, color = soil_moisture)) +
  geom_point(alpha = 0.6, size = 3) +
  facet_wrap(~gas, 
             scales = "free",
             labeller = as_labeller(c(
    co2 = "CO[2]",
    ch4 = "CH[4]",
    n2o = "N[2]*O"
  ), label_parsed)) +
  theme_classic() +
  labs(y = expression(atop("GHG Flux", "("*mu*"mol CO"[2]*" m"^{-2}*" s"^{-1}*"), (nmol CH"[4]*" / N"[2]*"O m"^{-2}*" s"^{-1}*")")),
       x = expression("Soil Temp. (C)"),
       fill = "Cover") +
  scale_fill_viridis_c() %>% 
  theme(
    strip.background = element_blank(),
    strip.text = element_text(
    size = 12,
    face = "bold"
  ))
ggsave("paper-figures/figS1.png", height = 3, width = 8)
```

# Use 3D plot -- only for exploration

```{r 3d-plot}
# library(dplyr)
# library(forcats)
# library(plotly)
# data_3d <- data %>% 
#   mutate(gas = fct_relevel(gas, "co2", "ch4", "n2o"))
# 
# plot_ly(
#   data = data_3d %>% filter(gas == "co2" & cover == "Wd"),
#   x = ~soil_temperature,
#   y = ~soil_moisture,
#   z = ~flux,
#   color = ~factor(date),
#   symbol = ~factor(date),
#   type = "scatter3d",
#   mode = "markers",
#   marker = list(size = 7, opacity = 0.7)
# ) %>%
#   layout(
#     scene = list(
#       xaxis = list(title = "Soil temperature (°C)"),
#       yaxis = list(title = "Soil moisture"),
#       zaxis = list(title = "GHG flux")
#     )
#   )
```

```{r co2-decision-tree}
d <- data %>%
  filter(gas == "co2") %>%
  mutate(
    cover = factor(cover),
    site = factor(site)
  ) %>%
  select(flux, soil_temperature, soil_moisture, cover) %>%
  na.omit()
tree_mod <- rpart(
  flux ~ soil_temperature + soil_moisture + cover,
  data = d,
  method = "anova",
  control = rpart.control(
    cp = 0.01,
    minsplit = 10,
    minbucket = 5
  )
)

# summary(tree_mod)
```

```{r co2-tree-eval}
# get best error
best_cp <- tree_mod$cptable[
  which.min(tree_mod$cptable[, "xerror"]),
  "CP"
]

best_cp

# prune tree
tree_pruned <- prune(tree_mod, cp = best_cp)

rpart.plot(tree_pruned)

printcp(tree_pruned)

# Training R²
pred <- predict(tree_pruned)
R2_train <- cor(d$flux, pred)^2

# Cross-validated R²
R2_cv <- 1 - min(tree_pruned$cptable[, "xerror"])

R2_train; R2_cv

tree_pruned$variable.importance/244.0486
```
nsplits = 6
CO2 CV R2 = 1-0.28630 = 0.71
Rel. Err. = 0.18

```{r ch4-decision-tree}
d <- data %>%
  filter(gas == "ch4") %>%
  mutate(
    cover = factor(cover),
    site = factor(site)
  ) %>%
  select(flux, soil_temperature, soil_moisture, cover) %>%
  na.omit()
tree_mod <- rpart(
  flux ~ soil_temperature + soil_moisture + cover,
  data = d,
  method = "anova",
  control = rpart.control(
    cp = 0.01,
    minsplit = 10,
    minbucket = 5
  )
)

# summary(tree_mod)
```

## CO2 Tree Q10

```{r}
df <- tibble(
  grouping = c("10-15", "15-20","15-20", "20-25", "20-25"),
  co2_flux = c(1.1, 1.8, 2.5, 2.8, 4.1),
  data_percent = c(0.3, 0.23, 0.16, 0.08, 0.22)
) %>% 
  group_by(grouping) %>% 
  # rowwise() %>% 
  mutate(group_percent = sum(data_percent),
         group_wt = data_percent/group_percent,
         group_wt_co2 = group_wt*co2_flux)  %>% 
  summarize(group_wt_co2 = sum(group_wt_co2))

# Add midpoints
df <- df %>%
  mutate(
    T_mid = case_when(
      grouping == "10-15" ~ 12.5,
      grouping == "15-20" ~ 17.5,
      grouping == "20-25" ~ 22.5
    )
  )

# 1. Interval-specific Q10
Q10_lower <- (df$group_wt_co2[2] / df$group_wt_co2[1]) ^ 2
Q10_upper <- (df$group_wt_co2[3] / df$group_wt_co2[2]) ^ 2

# 2. Overall Q10 (or regression)
Q10_overall <- df$group_wt_co2[3] / df$group_wt_co2[1]

fit <- lm(log(group_wt_co2) ~ T_mid, data = df)
Q10_fit <- exp(10 * coef(fit)["T_mid"])

# Results
tibble(
  interval = c("10-15 → 15-20", "15-20 → 20-25", "Overall / fitted (12.5–22.5 °C)"),
  Q10 = round(c(Q10_lower, Q10_upper, Q10_overall), 2)
)
```


```{r ch4-eval-tree}
# get best error
best_cp <- tree_mod$cptable[
  which.min(tree_mod$cptable[, "xerror"]),
  "CP"
]

best_cp

# prune tree
tree_pruned <- prune(tree_mod, cp = best_cp)

rpart.plot(tree_pruned)
printcp(tree_pruned)

# Training R²
pred <- predict(tree_pruned)
R2_train <- cor(d$flux, pred)^2

# Cross-validated R²
R2_cv <- 1 - min(tree_pruned$cptable[, "xerror"])

R2_train; R2_cv

tree_pruned$variable.importance/45.840316
```

```{r n2o-decision-tree}
d <- data %>%
  filter(gas == "n2o") %>%
  mutate(
    cover = factor(cover),
    site = factor(site)
  ) %>%
  select(flux, soil_temperature, soil_moisture, cover) %>%
  na.omit()
tree_mod <- rpart(
  flux ~ soil_temperature + soil_moisture + cover,
  data = d,
  method = "anova",
  control = rpart.control(
    cp = 0.01,
    minsplit = 10,
    minbucket = 5
  )
)

summary(tree_mod)

# get best error
best_cp <- tree_mod$cptable[
  which.min(tree_mod$cptable[, "xerror"]),
  "CP"
]

best_cp

# prune tree
tree_pruned <- prune(tree_mod, cp = best_cp)

rpart.plot(tree_pruned)
printcp(tree_pruned)

# Training R²
pred <- predict(tree_pruned)
R2_train <- cor(d$flux, pred)^2

# Cross-validated R²
R2_cv <- 1 - min(tree_pruned$cptable[, "xerror"])

R2_train; R2_cv

tree_pruned$variable.importance
```

## Comparison to Koyama data

```{r}
koyama <- read_csv("/Users/gawainmacneacail/Documents/Research/Work/Data Analysis/nachusa-2022-ghg/data/koyama/Koyama_et_al_GPM_Dryad_230720.csv") %>% 
  janitor::clean_names() %>% 
  mutate(ch4_mg_ch4_c_m_2_day = as.numeric(ch4_mg_ch4_c_m_2_day))

koyama %>% 
  ggplot(aes(x = ch4_mg_ch4_c_m_2_day)) +
  geom_density(fill = "red") +
  theme_bw()

  
koyama %>% 
  group_by(plot_id) %>% 
  summarize(median_ch4_mg_ch4_c_m_2_day = median(ch4_mg_ch4_c_m_2_day, na.rm=T)) %>% 
  arrange(median_ch4_mg_ch4_c_m_2_day) %>% 
  mutate(median_ch4_g_ch4_c_m_2_yr = median_ch4_mg_ch4_c_m_2_day*365.25/1000,
         median_ch4_kg_ch4_c_m_2_yr = median_ch4_g_ch4_c_m_2_yr*10)
```

